OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-16 paper/paper.Rmd
Fill in rest of numbers

**TODO: ref clustering method paragraph here.**
We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by
repeated 100 times.

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality (`r mcc_open_fit_db_vs_clust`% difference in median
MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
and reference databases.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
and `r open_vsearch_mcc` (respectively).
However, OptiClust and VSEARCH each ran `r sec_opticlust_vs_open_fit_db`% and 
`r sec_vsearch_vs_open_fit_db`% faster than OptiFit in open reference mode.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed only 
`r sec_opticlust_vs_vsearch`% as fast as VSEARCH.

In closed reference mode against databases, OptiFit produced lower quality OTUs
than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
dropped to `r closed_fit_rdp_mcc`.
An average of only `r frac_fit_db` of query sequences were fit to reference
database OTUs in closed reference mode across all dataset/database combinations.
VSEARCH was able to map `r frac_vsearch_vs_fit` more query sequences than
OptiFit to the Greengenes reference database.
In terms of run time, closed reference OptiFit outperformed OptiClust by 
`r sec_closed_fit_db_vs_clust` and VSEARCH by `r sec_closed_fit_db_vs_vsearch`.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (**TODO: diff in MCC medians**).
OTU quality was remarkably stable across 100 different random seeds (**TODO:
coeff of variation?**).
In terms of runtime, closed reference OptiFit performed faster than OptiClust on
whole datasets by **Z%**.
In open reference mode, OptiClust performed **X to Y%** faster than OptiFit only
when the OptiFit reference fraction was 30% or less.
The split dataset strategy performed just as well as the database strategy in
open reference mode regardless of database used, and outperformed the database
strategy in closed reference mode by **W%**.

We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs **X** and **Y** respectively), but **Z%** worse with similarity-weighted
sampling.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs decreases as the reference fraction increases from an MCC of
**A** with **J** reference sequences to an MCC of **B** with **K** reference
sequences.

@@

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to fit more sequence into OTUs than OptiFit in closed
reference mode.
While both mothur and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
mothur.
The OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use VSEARCH if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

>>>>

**TODO: ref clustering method paragraph here.**
We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by...
repeated 100 times.

In open reference mode, fitting the datasets to reference OTUs with OptiFit
produced OTUs of similar quality (`r mcc_open_fit_db_vs_clust`% difference in median
MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
and reference databases.
OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
and `r open_vsearch_mcc` (respectively).
OptiFit ran faster than VSEARCH by `r sec_vsearch_vs_open_fit_db`%  in open
reference mode, but slower than OptiClust by `r sec_opticlust_vs_open_fit_db`%.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.

In closed reference mode against databases, OptiFit produced lower quality OTUs
than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
dropped to `r closed_fit_rdp_mcc`.
An average of only `r frac_fit_db`% of query sequences were fit to reference
database OTUs in closed reference mode across all dataset/database combinations.
VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
OptiFit to the Greengenes reference database.
In terms of run time, closed reference OptiFit outperformed OptiClust by 
`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets (Fig. **XX**).
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
in open reference mode.

We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but worse for similarity-weighted
sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs decreases as the reference fraction increases from 
`r frac_fit_split_0.1`% of query sequences fit with `10`% of the dataset as the
reference, to `r frac_fit_split_0.8`% of query sequences fit with `80`% of the
dataset as the reference.

@@

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to map more sequences to OTUs than OptiFit in closed
reference mode.
While both mothur and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
mothur.
The OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use VSEARCH if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.
